Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-25 submission/manuscript.Rmd
clarify z-score in figure legend
ilakkiya edits and calculate tissue degs

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs) with unknown mechanisms. Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) that is mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene repression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a break against runaway activation of germline developmental programs in somatic lineages.

@@

To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes important for germ cell development, given the lack of a comprehensive list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g. meiotic versus spermatid differentiation genes.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated a list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@

@@

We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). An example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A.} Total number of germline-enriched RNA-seq DEGs for male hemizygous \textit{Kdm5c} knockout EpiLCs (XY 5CKO, purple), female heterozygous \textit{Kdm5c} knockout (XX 5CHET, orange), female homozygous \textit{Kdm5c} knockout (XX 5CKO, light orange) EpiLCs. \textbf{B.} Left - Eulerr overlap of \textit{Kdm5c} mutant male and female EpiLC germline DEGs. Right - Example of germline DEGs unique to females or males, \textit{Tktl2} and \textit{Esx1}.\textbf{C.} enrichPlot gene ontology analysis comparing enriched biological processes for germline DEGs shared between \textit{Kdm5c} mutant males and females (Shared), or unique to one sex (XX only or XY only). \textbf{D.} Heatmap of germline DEGs shared between male and female mutants. Color is the average log 2 fold change from sex-matched wild-type. \textbf{E.} Number of genes within each cluster from D. Clusters with higher expression in females compared to males (XX-biased) highlighted in orange. \textbf{F.}  UCSC browser view of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants (Average, n = 3). \textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome. Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant. X chromosome highlighted in black. \textbf{H.} Sankey diagram classifying egg-biased (pink) and sperm-biased (blue) and unbiased (purple) mouse germline-enriched genes. \textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants. \textbf{J.} UCSC browser view of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) germline genes dysregulated in both male and female \textit{Kdm5c} mutants (Average of n = 3).}
  \label{figurelabel}
\end{figure}

>>>>

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs) with unknown mechanisms. Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we examine how lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3) that is mutated in Claes-Jensen X-linked intellectual disability, contributes to tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a break against runaway activation of germline developmental programs in somatic lineages.

@@

To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Pathogenic mutations in the X chromosome gene _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, social deficits, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017; @bonefasSexuallyDimorphicAlterations2023]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone H3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in post-implantation epiblast cells[@borgelTargetsDynamicsPromoter2010; @velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Additionally, studies on germline gene repression have primarily been conducted in males and focused on select marker genes important for early germ cell development, given the lack of a comprehensive list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for different classes of germline genes, e.g. meiotic versus spermatid differentiation genes.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated a list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time. To illuminate the impact of sex upon germline gene suppression, we characterized germline transcripts expressed in male and female _Kdm5c_ mutants. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@

amy_res <- read.csv("../data/processed/TissueDEGs_amy5CKO.csv", header = TRUE)
amy_up = nrow(amy_res[amy_res$Direction == "Up",])
amy_tissue = nrow(amy_res[amy_res$Direction == "Up" & amy_res$tissue %in% tissue_res$tissue,])

hip_res <- read.csv("../data/processed/TissueDEGs_hip5CKO.csv", header = TRUE)
hip_up = nrow(hip_res[hip_res$Direction == "Up",])
hip_tissue = nrow(hip_res[hip_res$Direction == "Up" & hip_res$tissue %in% tissue_res$tissue,])

@@

We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are non-brain, tissue-specific genes (Amygdala: `r paste0(amy_tissue, "/", amy_up)` up DEGs, `r signif(amy_tissue/amy_up *100, 4)`% ; Hippocampus: `r paste0(hip_tissue, "/", hip_up)` up DEGs, `r signif(hip_tissue/hip_up *100, 4)`%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). An example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A.} Total number of germline-enriched RNA-seq DEGs for male hemizygous \textit{Kdm5c} knockout EpiLCs (XY 5CKO, purple), female heterozygous \textit{Kdm5c} knockout (XX 5CHET, orange), female homozygous \textit{Kdm5c} knockout (XX 5CKO, light orange) EpiLCs. \textbf{B.} Left - Eulerr overlap of \textit{Kdm5c} mutant male and female EpiLC germline DEGs. Right - Example of germline DEGs unique to females or males, \textit{Tktl2} and \textit{Esx1}.\textbf{C.} enrichPlot gene ontology analysis comparing enriched biological processes for germline DEGs shared between \textit{Kdm5c} mutant males and females (Shared), or unique to one sex (XX only or XY only). \textbf{D.} Heatmap of germline DEGs shared between male and female mutants. Color is the average log 2 fold change from sex-matched wild-type, z-scored across rows. \textbf{E.} Number of genes within each cluster from D. Clusters with higher expression in females compared to males (XX-biased) highlighted in orange. \textbf{F.}  UCSC browser view of a male and female shared germline DEG \textit{D1Pas1} that is more highly expressed in female mutants (Average, n = 3). \textbf{G.} Left - Number of all female germline DEGs located on each chromosome over the total number of germline-enriched genes on that chromsome. Right - Percentage of germline DEGs that lie on each chromosome for each \textit{Kdm5c} mutant. X chromosome highlighted in black. \textbf{H.} Sankey diagram classifying egg-biased (pink) and sperm-biased (blue) and unbiased (purple) mouse germline-enriched genes. \textbf{I.} Number of egg, sperm, or unbiased germline DEGs for male and female \textit{Kdm5c} mutants. \textbf{J.} UCSC browser view of egg-biased (\textit{Zp3}), sperm-biased (\textit{Dkkl1}), and unbiased (\textit{Dazl}) germline genes dysregulated in both male and female \textit{Kdm5c} mutants (Average of n = 3).}
  \label{figurelabel}
\end{figure}
